---
title: "Seminář - týden 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```


## tidyverse practice

Let's revisit `ukol_00_empty.Rmd`.

## Statistical Golems

- Why a "golem"? What similarity do we find?

- https://www.autodesk.com/research/publications/same-stats-different-graphs
  - inspired by Anscome's Quartet (70')

```{r}
# datasaurus_dozen imported
load(url("https://github.com/lockedata/datasauRus/blob/master/data/datasaurus_dozen.rda?raw=true"))

datasaurus_dozen %>% print()
datasaurus_dozen %>% count(dataset)

datasaurus_dozen %>% 
  ggplot(aes(x = x, y = y, colour = dataset)) + 
  geom_point() + facet_wrap(~dataset)
```

### Task

Check means and SD of x and y for each dataset. Check the correlations between x and y for each dataset.

```{r}



```

Show distributions of x and for each dataset

```{r}




```

## Small World, Large Worlds

- small world - optimality
- large world

## Garden of Forking Data

- https://archive.org/details/TheGardenOfForkingPathsJorgeLuisBorges1941/page/n7/mode/2up

### Example

- bag with 4 marbles (blue/white)
- goal: based on few draws, what is in the bag
- observation = BWB
- conjectures (CZ: domněnka)
- plausibility = normalized counting
- probability = set of non-negative numbers, sums to 1

```{r}
conjectures <- c("WWWW", "BWWW", "BBWW", "BBBW", "BBBB")
ways_to_produce <- function(conjs, observation) {
  result <- stringr::str_count(conjs, observation)
  names(result) <- conjs
  result
}

# how it works: single draw, one blue
ways_to_produce(conjectures, "B")
```

All the ways leading to BWB observation

```{r}
draw123 <- 
  ways_to_produce(conjectures, "B") *
  ways_to_produce(conjectures, "W") *
  ways_to_produce(conjectures, "B")
draw123
# can you make it into a table with conjectures?
```

One additional marble, blue again

```{r}
draw1234 <- draw123 *
  ways_to_produce(conjectures, "B")
draw1234
```

Extra info from factory: blue rare but always present

```{r}
factory_prior <- c(0, 3, 2, 1, 0)
draw1234 * factory_prior
```


### Terminology from textbook

- A conjectured proportion of blue marbles, p, is usually called a **parameter** value. It’s just a way of indexing possible explanations of the data.
- The relative number of ways that a value p can produce the data is usually called a **likelihood**. It is derived by enumerating all the possible data sequences that could have happened and then eliminating those sequences inconsistent with the data.
- The prior plausibility of any specific p is usually called the **prior probability**.
- The new, updated plausibility of any specific p is usually called the **posterior probability**.

### Example 2

- Same as Example 1 (4 draws), but different factory priors
- The bag may come from Factory A or Factory B (50:50 chance)
  - Factory A - same priors as in example (`c(0, 3, 2, 1, 0)`)
  - Factory B - likely to produce pairs (`c(0, 1, 2, 1, 0)`)

```{r}
factory_prior_a <- c(0, 3, 2, 1, 0)
factory_prior_b <- c(0, 1, 2, 1, 0)





```

What of Factory B produces also uniform bags? (`c(2, 1, 2, 1, 2)`)

```{r}
factory_prior_b2 <- c(2, 1, 2, 1, 2)



```

### Example 3

Imagine new set of marbles:

- four marbles again, three colours (red, blue, white)
- you draw 3 times: RBW
- one more draw: R

Build your conjectures

- [combination with repetition](https://en.wikipedia.org/wiki/Combination#Number_of_combinations_with_repetition)
- you can count them with `choose(n + k - 1, k)`, *n* number of colours, *k* number of draws

```{r}
conjectures <- c(
  
  
)
choose(4 + 3 - 1, 4)
```

All the ways leading to RBW

```{r}
# d123 <- 

d123
```

One more draw: R

```{r}


```



