---
title: "Seminář - týden 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(patchwork)
library(brms)
library(rethinking) # for Quadratic approx + HW
```

## Examples 2 and 3 about marbles

Revisiting exercises from last week.

## Tossing example

- what does it mean: Design > Condition > Evaluate
- Design = data story
  - How do the data arise?
  - Translate data story into probability statements
- Condition = update
  - Bayesian updating defines optimal learning in small world, converts *prior* into *posterior*
  - Give your golem an information state, before the data: Here, an initial confidence in each possible value of p between zero and one
  - Condition on data to update information state: New confidence in each value of p, conditional on data
  

- *p* = true proportion of *water*
- toss data: W L W W W L W L W
- tosses are independent


```{r}
# recreating the example
# 1) data
d <- tibble(toss = c("w", "l", "w", "w", "w", "l", "w", "l", "w"))
d <- d %>% 
  mutate(n_trials  = 1:9, n_success = cumsum(toss == "w"))
```

```{r}
sequence_length <- 50

dd <- # ** calculate curves
  d %>% 
  expand(
    nesting(n_trials, toss, n_success), 
    p_water = seq(from = 0, to = 1, length.out = sequence_length)) %>% 
  group_by(p_water) %>% 
  # ** lagging required for dashed curves (priors)
  # you can learn more about lagging here: https://www.rdocumentation.org/packages/stats/versions/3.5.1/topics/lag or here: https://dplyr.tidyverse.org/reference/lead-lag.html
  mutate(lagged_n_trials  = lag(n_trials,  k = 1),
         lagged_n_success = lag(n_success, k = 1)) %>% 
  ungroup() %>% 
  # ** this makes the curves (priors, likelihood)
  mutate(prior      = ifelse(n_trials == 1, .5,
                             dbinom(x    = lagged_n_success, 
                                    size = lagged_n_trials, 
                                    prob = p_water)),
         likelihood = dbinom(x    = n_success, 
                             size = n_trials, 
                             prob = p_water),
         strip      = str_c("n = ", n_trials)
         ) %>% 
  # the next three lines allow us to normalize the prior and the likelihood, 
  # putting them both in a probability metric 
  group_by(n_trials) %>% 
  mutate(prior      = prior      / sum(prior),
         likelihood = likelihood / sum(likelihood)) %>%
  identity() # ** dummy code
  
  # plot!
  dd %>% 
  ggplot(aes(x = p_water)) +
  geom_line(aes(y = prior), linetype = 2) +
  geom_line(aes(y = likelihood)) +
  scale_x_continuous("proportion water", breaks = c(0, .5, 1)) +
  #scale_y_continuous("plausibility", breaks = NULL) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~strip, scales = "free_y") +
  NULL # ** dummy code
```

## Calculations with `dbinom`

```{r}
# check ?dbinom
# think about difference:
dbinom(6, 10, p = 0.7)
dbinom(6:7, 10, p = 0.7)
# vs
p10 <- (0:10)/10
dbinom(6, 10, p = p10)

# in pictures
qplot(0:10, dbinom(0:10, 10, p = 0.7)) + qplot(p10, dbinom(6, 10, p = p10))

# check sums
sum(dbinom(0:10, 10, p = 0.7))
sum(dbinom(6, 10, p = p10))
```

## Introducing `dbeta`

- Beta distribution
- Some insights https://towardsdatascience.com/beta-distribution-intuition-examples-and-derivation-cf00f4db57af
- probability of probability
- example: we see 6 heads + 4 tails

```{r}
p100 <- (0:100)/100
qplot(p100, dbeta(p100, 6 + 1, 4 + 1))
# try changing "6" and "4" ... 
qplot(p100, dbeta(p100, 0 + 1, 0 + 1))
qplot(p100, dbeta(p100, 1 + 1, 0 + 1))
qplot(p100, dbeta(p100, 1 + 1, 1 + 1))
qplot(p100, dbeta(p100, 2 + 1, 1 + 1))
qplot(p100, dbeta(p100, 2 * 5 + 1, 1 * 5 + 1))
qplot(p100, dbeta(p100, 2 * 25 + 1, 1 * 25 + 1))
qplot(p100, dbeta(p100, 2 * 125 + 1, 1 * 125 + 1))

# dbinom is normalised probability density
qplot(dbeta(p100, 6 + 1, 4 + 1), dbinom(6, 10, p = p100))
```

# Three motors

## Grid approximation

(1) Define the grid. This means you decide how many points to use in estimating the posterior, and then you make a list of the parameter values on the grid.
(2) Compute the value of the prior at each parameter value on the grid.
(3) Compute the likelihood at each parameter value.
(4) Compute the unstandardized posterior at each parameter value, by multiplying the prior by the likelihood.
(5) Finally, standardize the posterior, by dividing each value by the sum of all values.


```{r}
d <-
  tibble(
    # define grid
    p_grid = seq(from = 0, to = 1, length.out = 20), 
    # define prior
    prior = 1
  ) %>% 
  mutate(
    # compute likelihood at each value in grid
    likelihood = dbinom(6, size = 9, prob = p_grid)) %>% 
  # compute product of likelihood and prior
  mutate(unstd_posterior = likelihood * prior) %>% 
  # standardize the posterior, so it sums to 1
  mutate(posterior = unstd_posterior / sum(unstd_posterior)) 
```

```{r}
d %>% 
  ggplot(aes(x = p_grid, y = posterior)) +
  geom_point() +
  geom_line() +
  labs(subtitle = "20 points",
       x = "probability of water",
       y = "posterior probability") +
  theme(panel.grid = element_blank())
```

## Quadratic approxiamtion

```{r}
globe_qa <-
  rethinking::map(
    alist(
      w ~ dbinom(36, p),  # binomial likelihood
      p ~ dunif(0, 1)    # uniform prior
    ), 
    data = list(w = 24))

# display summary of quadratic approximation
precis(globe_qa)
precis(globe_qa, prob = .95)
```

## Markov chain Monte Carlo

```{r message=FALSE, warning=FALSE, results="hide"}
globe_qa_brms <-
  brm(data = list(w = 24), 
      family = binomial(link = "identity"),
      w | trials(36) ~ 1,
      prior(beta(1, 1), class = Intercept),
      iter = 4000, warmup = 1000,
      control = list(adapt_delta = .9),
      seed = 4)
```

```{r}
print(globe_qa_brms)
```

```{r}
posterior_samples(globe_qa_brms) %>% 
  mutate(n = "n = 36") %>%

  ggplot(aes(x = b_Intercept)) +
  geom_density(fill = "black") +
  labs(x = "proportion water") +
  xlim(0, 1) +
  theme(panel.grid = element_blank()) +
  facet_wrap(~n)
```

# Book questions

## Book question 2E1

Which of the expressions below correspond to the statement: the probability of rain on Monday?

(1) Pr(rain)
(2) Pr(rain|Monday)
(3) Pr(Monday|rain)
(4) Pr(rain,Monday)/Pr(Monday)

```{r}
# definition of conditional
# p(A and B) = p(A | B) * p(B)
# Bayes theorem
# p(A | B) = p(B | A) * p(A) / p(B)
```


## Book question 2E2

Which of the following statements corresponds to the expression: Pr(Monday|rain)?

(1) The probability of rain on Monday.
(2) The probability of rain, given that it is Monday.
(3) The probability that it is Monday, given that it is raining. 
(4) The probability that it is Monday and that it is raining.

## Book question 2E3

Which of the expressions below correspond to the statement: the probability that it is Monday, given that it is raining?

(1) Pr(Monday|rain)
(2) Pr(rain|Monday)
(3) Pr(rain|Monday) * Pr(Monday)
(4) Pr(rain|Monday) * Pr(Monday) / Pr(rain) 
(5) Pr(Monday|rain) * Pr(rain) / Pr(Monday)

## Book question 2E4

The Bayesian statistician Bruno de Finetti (1906–1985) began his 1973 book on probability theory with the declaration: "PROBABILITY DOES NOT EXIST." The capitals appeared in the original, so I imagine de Finetti wanted us to shout this statement. What he meant is that probability is a device for describing uncertainty from the perspective of an observer with limited knowledge; it has no objective reality. Discuss the globe tossing example from the chapter, in light of this statement. What does it mean to say "the probability of water is 0.7"?

## Book question 2M1

Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p.

(1) W,W,W
(2) W,W,W,L
(3) L,W,W,L,W,W,W

## Book question 2M2

Now assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when p ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of observations in the problem just above.

## Book question 2M3

Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 70% covered in water. The Mars globe is 100% land. Further suppose that one of these globes—you don’t know which—was tossed in the air and produced a “land” observation. Assume that each globe was equally likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on seeing “land” (Pr(Earth|land)), is 0.23.

## Book question 2M4

Suppose you have a deck with only three cards. Each card has two sides, and each side is either black or white. One card has two black sides. The second card has one black and one white side. The third card has two white sides. Now suppose all three cards are placed in a bag and shuffled. Someone reaches into the bag and pulls out a card and places it flat on a table. A black side is shown facing up, but you don’t know the color of the side facing down. Show that the probability that the other side is also black is 2/3. Use the counting method (Section 2 of the chapter) to approach this problem. This means counting up the ways that each card could produce the observed data (a black side facing up on the table).

## Book question 2M5

Now suppose there are four cards: B/B, B/W, W/W, and another B/B. Again suppose a card is drawn from the bag and a black side appears face up. Again calculate the probability that the other side is black.

## Book question 2M6

Imagine that black ink is heavy, and so cards with black sides are heavier than cards with white sides. As a result, it’s less likely that a card with black sides is pulled from the bag. So again assume there are three cards: B/B, B/W, and W/W. After experimenting a number of times, you conclude that for every way to pull the B/B card from the bag, there are 2 ways to pull the B/W card and 3 ways to pull the W/W card. Again suppose that a card is pulled and a black side appears face up. Show that the probability the other side is black is now 0.5. Use the counting method, as before.

## Book question 2M7

Assume again the original card problem, with a single card showing a black side face up. Before looking at the other side, we draw another card from the bag and lay it face up on the table. The face that is shown on the new card is white. Show that the probability that the first card, the one showing a black side, has black on its other side is now 0.75. Use the counting method, if you can. *Hint: Treat this like the sequence of globe tosses, counting all the ways to see each observation, for each possible first card.*

## Book question 2H1

Suppose there are two species of panda bear. Both are equally common in the wild and live in the same places. They look exactly alike and eat the same food, and there is yet no genetic assay capable of telling them apart. They differ however in their family sizes. Species A gives birth to twins 10% of the time, otherwise birthing a single infant. Species B births twins 20% of the time, otherwise birthing singleton infants. Assume these numbers are known with certainty, from many years of field research.
Now suppose you are managing a captive panda breeding program. You have a new female panda of unknown species, and she has just given birth to twins. What is the probability that her next birth will also be twins?

## Book question 2H2

Recall all the facts from the problem above. Now compute the probability that the panda we have is from species A, assuming we have observed only the first birth and that it was twins.

## Book question 2H3 

Continuing on from the previous problem, suppose the same panda mother has a second birth and that it is not twins, but a singleton infant. Compute the posterior probability that this panda is species A.

## Book question 2H4

A common boast of Bayesian statisticians is that Bayesian inference makes it easy to use all of the data, even if the data are of different types.

So suppose now that a veterinarian comes along who has a new genetic test that she claims can identify the species of our mother panda. But the test, like all tests, is imperfect. This is the information you have about the test:

- The probability it correctly identifies a species A panda is 0.8
- The probability it correctly identifies a species B panda is 0.65

The vet administers the test to your panda and tells you that the test is positive for species A. First ignore your previous information from the births and compute the posterior probability that your panda is species A. Then redo your calculation, now using the birth data as well.

